---
title: "Labour-Market Tightness and Slack in Australia"
author: "Rimlan & Trisha"
format: html
bibliography: references.bib
execute:
  working-dir: ..
---

**Executive Summary**

To be completed after the analyis and discussion are finalised.

# 1. Introduction 

## 1.1 Background
The labour market plays an important role in the Australian economy by connecting employers seeking workers with individuals who are available and willing to work. Labour-market conditions change over time as the demand for workers and the availability of labour respond to economic conditions. Understanding these changes requires consideration of both the demand for labour and the extent to which available labour resources are being utilised.

One indicator of labour demand is the number of job vacancies. A higher level of vacancies generally indicates greater demand from employers for additional workers. However, different vacancy measures capture different aspects of recruitment activity. The Australian Bureau of Statistics (ABS) measures job vacancies as jobs that are available for immediate filling on a specific reference date and for which recruitment action has been taken [@ABS2026JobVacancies]. As a result, ABS Job Vacancies represents a stock measure of vacancies at a particular point in time.

Jobs and Skills Australia (JSA) provides another measure of labour demand through the Internet Vacancy Index (IVI). The IVI is a monthly series based on newly lodged online job advertisements from SEEK, CareerOne and Workforce Australia. It provides detailed information by occupation, skill level and geographic area and can be used as an indicator of recruitment activity and labour demand. However, because it measures newly lodged advertisements during each month rather than vacancies existing at a particular point in time, it differs conceptually from the ABS Job Vacancies measure.

Labour demand alone does not provide a complete picture of labour-market conditions. The supply side of the labour market is also important, particularly the amount of labour that is available but not fully utilised. This is commonly described as labour-market slack or labour underutilisation. Unemployment is one measure of unused labour capacity, but it does not capture workers who already have employment but would prefer and are available to work additional hours. The ABS therefore also publishes measures of underemployment and underutilisation. Underutilisation combines unemployment and underemployment and provides a broader measure of unused labour capacity in the labour market.

Considering labour demand and labour-market slack together can provide a more complete view of labour-market conditions. For example, a period with relatively high demand for workers and limited available labour represents different conditions from a period with weak labour demand and substantial unused labour capacity. This motivates examining not only vacancies and slack separately, but also how they move in relation to one another over time.

A related concept is labour-market tightness, which compares labour demand with the availability of workers. In this project, an initial measure of tightness is constructed as the ratio of ABS Job Vacancies to unemployed persons. A higher vacancy-to-unemployment ratio represents more vacancies relative to unemployed workers, while a lower ratio represents fewer vacancies relative to the number of unemployed workers. This measure provides an additional perspective on the balance between labour demand and available labour.

## 1.2 Research Motivation
Job vacancies, unemployment, underemployment and underutilisation each provide useful information about the Australian labour market, but they measure different aspects of labour-market conditions. Examining any one indicator independently may therefore provide an incomplete picture.

This distinction is particularly relevant when measuring labour-market slack. Unemployment identifies people without employment who satisfy specific job-search and availability criteria, whereas underemployment captures employed people whose labour is not being fully utilised. The ABS defines underemployed workers as employed people who would prefer and are available to work more hours than they currently have. The underutilisation rate provides a broader indicator by combining unemployed and underemployed people relative to the labour force.

The measurement of labour demand also requires careful consideration. The IVI captures the flow of newly lodged online job advertisements, while ABS Job Vacancies measures the stock of vacancies available for immediate filling on a reference date. In addition, the IVI does not represent all recruitment activity in Australia because it only covers advertisements from its contributing online job boards and excludes recruitment through other websites and informal channels. These differences mean that the choice of vacancy measure depends on the purpose of the analysis.

This project is therefore motivated by the need to examine labour demand and labour-market slack jointly rather than treating them as independent indicators. In particular, examining changes in these measures allows the analysis to investigate whether increases or decreases in job vacancies are associated with corresponding changes in unused labour capacity.

Labour-market tightness is also considered as a complementary measure. By relating the stock of job vacancies to the number of unemployed persons, tightness provides information about vacancies relative to the available pool of unemployed workers. This allows the project to investigate both the direct relationship between changes in vacancies and slack and the broader relationship between labour-market tightness and different measures of slack.

## 1.3 Research Question
Based on this motivation, the main research question is:

To what extent are changes in job vacancies associated with changes in labour-market slack in Australia?

The wording of the research question focuses on association rather than causation. The purpose of the analysis is to determine whether movements in labour demand are systematically related to movements in labour-market slack and to examine how this relationship differs depending on the measure of slack considered.

The project considers three measures of labour-market slack: unemployment, underemployment and underutilisation. This allows the relationship to be examined using both a narrower measure of unused labour capacity and broader measures that account for workers who are employed but would prefer additional hours.

Labour-market tightness is examined as a complementary perspective rather than as a replacement for the main vacancy measure. This distinction is important because the research question specifically concerns changes in job vacancies, while tightness incorporates information from both vacancies and unemployment.

## 1.4 Research Objectives
The project has four main objectives:

To describe how job vacancies and labour-market slack have changed in Australia over the available data period.
To compare alternative measures of labour-market slack, including unemployment, underemployment and underutilisation.
To examine the association between changes in job vacancies and changes in labour-market slack, including potential lagged relationships.
To investigate labour-market tightness as an additional indicator of the balance between labour demand and available labour.

The analysis uses data from the Australian Bureau of Statistics and Jobs and Skills Australia. The IVI is used to examine monthly online recruitment activity, while ABS Job Vacancies and Labour Force data provide measures of vacancies and labour availability. The different data sources are examined carefully because their definitions, frequency and measurement concepts are not identical.

## 1.5 Report Structure
The remainder of this report is organised as follows. Section 2 introduces the main labour-market concepts and provides background on labour demand, labour-market slack and labour-market tightness. Section 3 describes the datasets, variable construction and measurement choices used in the analysis. Section 4 presents exploratory analysis of labour demand, slack and tightness in Australia. Section 5 describes the analytical methodology, including the analysis of changes, correlations and potential lagged relationships. Section 6 presents the empirical results, followed by a discussion of their interpretation in Section 7. Section 8 discusses the limitations of the analysis and possible extensions, and Section 9 concludes the report.

# 2. Labour-Market Concepts and Background

## 2.1 Labour Demand and Job Vacancies

## 2.2 Labour-Market Slack

Labour market slack refers to the extent to which available labour resources are not fully utilised. While unemployment captures people who are without employment and actively seeking and avaiable to work, it does not capture all forms o unused labour capacity. Some employed people may wish to work additional hours, while other potential forms of labour underutilisation may not be reflected in the unemployment rate alone. 

This project therefore considers three related measures from the Australian Bureau of Statistics (ABS) Labour Force Survey: unemployment, underemployment and underutilisation. Unemployment provides an indicator of the number of people who are not employed but are available for work and actively seeking employment. Underemployment captures employed people who would like and are available to work additional hours. Underutilisation combines unemployment and underemployment and therefore provides a broader measure of labour that is not being fully utilised. 

Underutilisation is used in this project as the primary observable proxy for labour market slack. This is because slack is not directly observable as a single variable and underutilisation captures both the absence of employment and insufficient hours among employed people. Unemployment and underemployment are also analysed separately to identify whether changes in overall slack are associated more strongly with changes in the number of people without work or with the availability of additional hours among employed workers. 

The use of underutilisation as a proxy does not imply that it represents all unused labour capacity in the economy. Labour market slack can include dimensions that are not fully captured by the ABS measures, such as people who are marginally attached to the labour force or workers whose available hours do not fully reflect their preferred working arrangements. The measures should therefore be interpreted as observable indicators of labour underutilisation rather than a complete measure of all potential labour resources. 

Examining labour market slack alongside labour demad provides a basis for assessing how changes in employer demand are associated with the availability of unused labour. In periods of stronger labour demand, slack may decline as more people move into employment or increase their hours. Conversely, weaker labour demand may be associated with higher unemployment or underemployment. The relationship is not necessarily one-to-one, however, because labour market outcomes can also be affected by participation, industry composition, working hours and broader economic conditions. 

## 2.3 Labour-Market Tightness

Labour market tightness describes the balance between the demand for workers and the availability of workers seeking employment. A labour market is generally considered tighter when employers have a relatively large number of vacancies compared with the number of unemployed people available to fill them. Conversely, a less tight labour market occurs when there are fewer vacancies relative to the number of unemployed people.

In this project, labour market tightness is measured using the vacancy to unemployment ratio. The measure is calculated as the number of job vacancies relative to the number of unemployed people:

$$
\text{Labour-Market Tightness} =
\frac{\text{Job Vacancies}}{\text{Unemployed Persons}}
$$

A higher ratio indicates that there are more vacancies relative to unemployed people, while a lower ratio indicates that unemployed people are more numerous relative to available vacancies. The measure therefore provides an indicator of the balance between labour demand and the available pool of unemployed workers.

The vacancy to unemployment ratio is useful because considering vacancies and unemployment separately does not fully describe the conditions faced by employers and job seekers. An increase in vacancies may indicate stronger labour demand, but its implications for labour market conditions depend partly on how many unemployed workers are available. Similarly, a decline in unemployment may indicate improving labour market conditions, but the degree of tightness also depends on the number of vacancies available.

In this analyis, the tightness measure is constructed using the seasonally adjusted ABS Job Vacancies series and the seasonally adjusted ABS number of unemployed persons. The resulting measure is examined at a quarterly frequency to investigate its relationship with labour market slack, particularly unemployment, underemployment and underutilisation. 

Labour market tightness and labour market slack represent two related but distinct dimensions of labour market conditions. Tightness focuses on the relationship between employer demand and the pool of unemployed workers, whereas slack focuses on the extent to which available labour in underutilised. Examining both measures together provides a broader view of whether changes in labour demand are associated with changes in unused labour capacity in Australia. 

## 2.4 Relationship Between Labour Demand and Slack

## 2.5 Australian Labour-Market Context

# 3. Data and Measurement

## 3.1 Jobs and Skills Australia Internet Vacancy Index

### 3.1.1 Data Source and Collection

The Internet Vacancy Index (IVI), published by Jobs and Skills Australia
(JSA), is used in this project as a monthly indicator of labour demand in
Australia. The IVI measures online job advertisements newly lodged during
each reference month on three contributing job boards: SEEK, CareerOne and
Workforce Australia [@JSAIVIMethodology].

The IVI is based on administrative data provided by the contributing job
boards rather than data collected through an employer survey. JSA processes
the advertisements and classifies them by occupation, skill level and
geographical location. Duplicate advertisements posted on the same or
multiple contributing job boards are removed during this process
[@JSAIVIMethodology].

### 3.1.2 Measurement and Coverage
The IVI is available as a monthly time series from January 2006. The data
provide information by occupation, skill level, state and territory, and
regional areas. For the national labour-demand analysis in this project,
the seasonally adjusted Australian total series is used.

An important characteristic of the IVI is that it measures newly lodged
advertisements during a month. It therefore represents a flow of online
recruitment activity rather than the stock of vacancies that remain
unfilled at a particular point in time [@JSAIVIMethodology].

### 3.1.3 Limitations of the IVI
The IVI does not represent all vacancies in the Australian labour market.
It only includes advertisements from the contributing online job boards
and therefore excludes recruitment through other job boards, employer
websites, social media, newspapers and informal methods such as word of
mouth. In addition, one advertisement may represent multiple available
positions, while some vacancies may be filled without being advertised
online [@JSAIVIMethodology].

For these reasons, the IVI is interpreted in this project as an indicator
of online recruitment activity and labour demand rather than a complete
measure of all job vacancies in Australia.


## 3.2 ABS Job Vacancies

### 3.2.1 Job Vacancies Survey
The second measure of labour demand is obtained from the Australian Bureau
of Statistics (ABS) Job Vacancies Survey (JVS). Unlike the IVI, which is
based on online job advertisements, the JVS collects information directly
from a sample of employing organisations. Approximately 4,000 statistical
units are selected from the ABS Business Register, with data collected
through online electronic collection and/or telephone interviews
[@ABS2026JobVacancies].

The ABS defines a job vacancy as a job that is available for immediate
filling on the survey reference date and for which recruitment action has
been taken [@ABS2026JobVacancies].

### 3.2.2 Measurement and Coverage
The JVS is conducted quarterly. Its reference date is the third Friday of
the middle month of each quarter, corresponding to February, May, August
and November. The number of vacancies therefore refers to vacancies
existing on a particular reference date rather than vacancies accumulated
over the entire quarter [@ABS2026JobVacancies].

For this project, the seasonally adjusted Australian total job-vacancy
series is used. ABS classifies job vacancies as a stock measure because
they are measured at a specific point in time [@ABS2026JobVacancies].

There are five missing quarterly observations between August 2008 and
August 2009 because the Job Vacancies Survey was suspended after May 2008
and reinstated in November 2009. These observations are retained as
missing rather than being imputed, as the ABS does not provide official
estimates for the gap period [@ABS2026JobVacancies].

### 3.2.3 IVI versus ABS Job Vacancies
Although both series provide information about labour demand, they measure
different concepts. The IVI counts newly lodged online job advertisements
during a month and is therefore a flow measure of recruitment activity.
In contrast, ABS Job Vacancies measures the number of vacancies available
for immediate filling on a particular reference date and is therefore a
stock measure [@JSAIVIMethodology; @ABS2026JobVacancies].

This distinction is important for the analysis. The IVI is used to examine
monthly changes in recruitment demand, while ABS Job Vacancies is used in
constructing the quarterly labour-market tightness measure. Using ABS Job
Vacancies for the tightness measure provides a more consistent comparison
with the number of unemployed persons, since both represent quantities
measured for a particular reference period.


## 3.3 ABS Labour Force Data
This section describes the ABS Labour Force data used to measure
unemployment, underemployment and underutilisation.

### 3.3.1 Data Source and Labour-Market Measures
The analysis of labour market slack uses data from the Australian Bureau of Statistics (ABS) Labour Force Survey. The Labour Force Survey provides monthly measures of employment, unemployment, undermeployment and labour-force participation for the Australian population aged 15 years and over. These measures provide the basis for constructing observable indicators of labour market slack over time. 

Three measures are central to this analysis. The unemployment rate captures the proportion of the labour force that is unemployed, while the underemployment rate captures employed people who want and are available to work additional hours. The underutilisation rate combines unemployment and underemployment and provides a broader measure of labour underutilisation. 

The analysis retains unemployment and underemployment as separate measures while using underutilisation as the primary proxy for overall labour market slack. This approach allows the analysis to distinguish between different forms of unused labour capacity rather than relying on unemployment alone. 

### 3.3.2 Measurement of Labour-Market Slack
Labour market slack is not directly observed as a single variable. Instead, it is approximated using observable measures of labour underutilisation. In this study, underutilisation is used as the main proxy because it incorporates both unemployment and underemployment. 

This distinction is important because unemployment alone may understate available labour capacity. An individual who is employed but wants and is available to work additional hours is not classified as unemployed, but their unused working time capacity is captured through underemployment. Including underemployment therefore provides a broader perspective on labour market slack. 

The analysis also examines unemployment and underemployment separately. This enables changes in overall underutilisation to be interpreted in terms of whether they are associated primarily with changes in unemployment, changes in the availability of additional working hours, or movements in both measures. 

### 3.3.3 Data Period and Frequency

The ABS Labour Force data used in this study contain 582 monthly observations covering the period from February 1978 to July 2026. The long time span allows labour market slack to be examined across different economic conditions and major periods of labour market disruption. 

For the historical analysis, the data are grouped into six broad periods: 1978-1989, 1990-1994, 1995-2007, 2008-2019, 2020-2021 and 2022 onwards. These periods provide a basis for comparing labour market conditions across different phases of the Australian economy, including the early 1990s recession and the COVID-19 period. 

The monthly frequency of the Labour Force Survey also allows changes in unemployment, underemployment and underutilisation to be examined at a relatively detailed time scale. For the analysis of the labour market tightness, the monthly labour force measures are aligned with the quarterly ABS Job Vacancies data to construct the vacancy to unemployment ratio. 

### 3.3.4 Measurement Considerations

The ABS Labour Force measures provide consistent and widely used indicators of labour market conditions, but they do not capture every dimension of unused labour capacity. In particular, underutilisation does not represent all potnetial forms of labour market slack, such as the capacity of people who are outside the labour force to enter employment or changes in workers preferred working arrangements that are not reflected in the standard measures. 

Underutilisation should thereore be interpreted as an observable proxy for labour market slack rather than a complete measure of unused labour resources. This limitation is important when interpreting the relationship between labour market tightness and slack. The analysis focuses on statistical associations between labour market tightness and slack. The analysis focuses on statistical associations between the measures and does not interpret these relationships as evidence of causality. 


## 3.4 Labour-Market Tightness

### 3.4.1 Vacancy-to-Unemployment Ratio
Labour-market tightness is used to represent the balance between labour
demand and the availability of unemployed workers. In this project,
tightness is initially measured using the vacancy-to-unemployment ratio:

$$
\text{Labour-market tightness}_t =
\frac{\text{Job Vacancies}_t}
{\text{Unemployed Persons}_t}
$$

A higher value indicates that there are more job vacancies relative to
the number of unemployed persons, representing a tighter labour market.
Conversely, a lower value indicates fewer vacancies relative to unemployed
persons.

### 3.4.2 Construction of the Measure
The numerator is the seasonally adjusted Australian total ABS Job
Vacancies series, while the denominator is the seasonally adjusted number
of unemployed persons from the ABS Labour Force data.

Because Job Vacancies data are quarterly while Labour Force data are
monthly, unemployed-person observations are matched to the corresponding
vacancy reference months of February, May, August and November. Both
series are measured in thousands of persons or positions, so the units
cancel when the ratio is calculated.

The resulting measure provides a quarterly series of labour-market
tightness that can subsequently be compared with the measures of
labour-market slack used in this project.

## 4.1 ABS Job Vacancies Over Time
The ABS Job Vacancies series provides a quarterly view of labour demand in Australia. Figure @fig-job-vacancies shows the seasonally adjusted number of job vacancies over the study period.

```{r}
#| label: fig-job-vacancies
#| fig-cap: "Seasonally adjusted job vacancies in Australia"
#| warning: false
#| message: false

library(tidyverse)

job_vacancies <- read_csv(
  "../data/processed/abs_job_vacancies_clean.csv",
  show_col_types = FALSE
) |>
  mutate(Quarter = as.Date(Quarter))

ggplot(job_vacancies, aes(x = Quarter, y = Job_Vacancies)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Job vacancies ('000)",
    title = "Job Vacancies in Australia",
    subtitle = "Seasonally adjusted ABS Job Vacancies"
  ) +
  theme_minimal()
```

Job vacancies changed considerably over the study period. Vacancies declined during the late 2000s, although the series contains a gap between August 2008 and August 2009 due to the suspension of the Job Vacancies Survey during this period [@ABS2026JobVacancies].

A particularly large decline is visible in 2020, followed by a strong increase during 2021 and 2022. Job vacancies reached their highest levels around 2022 before declining over the following years.

These patterns indicate that labour demand has varied substantially over time. However, the figure is descriptive and does not by itself establish the causes of these movements.

## 4.2 Internet Vacancy Index Over Time
The Internet Vacancy Index provides a higher-frequency view of recruitment activity because it records newly lodged online job advertisements each month. Figure @fig-ivi-total shows the seasonally adjusted Australian total IVI.

```{r}
#| label: fig-ivi-total
#| fig-cap: "Internet Vacancy Index for Australia"
#| warning: false
#| message: false

library(tidyverse)

ivi_anzsco2 <- read_csv(
  "../data/processed/ivi_anzsco2_clean.csv",
  show_col_types = FALSE
)

ivi_australia <- ivi_anzsco2 |>
  filter(
    Level == 1,
    ANZSCO_CODE == "0",
    State == "AUST"
  ) |>
  arrange(Month)

ggplot(ivi_australia, aes(x = Month, y = Vacancies)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Online job advertisements",
    title = "Internet Vacancy Index in Australia",
    subtitle = "Seasonally adjusted national series"
  ) +
  theme_minimal()
```

The IVI also shows substantial variation in recruitment activity over time. Online job advertisements increased during the earlier part of the series before declining around 2008–2009. Recruitment activity subsequently recovered, although further periods of decline and recovery are visible.

One of the most pronounced movements occurs in 2020, when the IVI falls sharply. This is followed by a strong increase, with recruitment activity reaching particularly high levels around 2022. The index subsequently declines through the following years.

The broad movements are similar to those observed in ABS Job Vacancies. However, the two series should not be interpreted as identical measures. The IVI measures the flow of newly lodged online advertisements each month, whereas ABS Job Vacancies measures the stock of vacancies available for immediate filling at a reference date [@JSAIVIMethodology; @ABS2026JobVacancies].

## 4.3 Vacancy Patterns by Occupation

While the national IVI shows changes in overall recruitment activity, labour demand is not evenly distributed across occupations. The broad occupation-level IVI data are therefore examined to compare recruitment activity across major occupational groups.

The latest available observation is used to provide a snapshot of the occupational composition of online recruitment demand.

```{r}
#| label: fig-occupation-vacancies
#| fig-cap: "Online job advertisements by broad occupation group"
#| warning: false
#| message: false

library(tidyverse)

latest_month <- max(ivi_anzsco2$Month, na.rm = TRUE)

occupation_latest <- ivi_anzsco2 |>
  filter(
    Level == 2,
    State == "AUST",
    Month == latest_month
  ) |>
  arrange(Vacancies) |>
  mutate(Title = factor(Title, levels = Title))

ggplot(
  occupation_latest,
  aes(x = Vacancies, y = Title)
) +
  geom_col() +
  labs(
    x = "Online job advertisements",
    y = NULL,
    title = "Online Job Advertisements by Occupation",
    subtitle = paste("Australia,", latest_month)
  ) +
  theme_minimal()
```

Figure @fig-occupation-vacancies shows that online recruitment demand differs substantially across broad occupational groups. Professionals recorded the highest number of online job advertisements in the latest observation, indicating particularly high recruitment activity within this broad occupational category.

Other occupational groups also account for substantial numbers of advertisements, demonstrating that labour demand is distributed across multiple parts of the Australian labour market. The differences across occupations are important because movements in aggregate vacancy measures may reflect different recruitment conditions across occupational groups.

These results describe the composition of online recruitment activity rather than the total number of vacancies in each occupation. As discussed in Section 3, the IVI only captures newly lodged advertisements from the contributing online job boards and should therefore be interpreted as an indicator of recruitment activity [@JSAIVIMethodology].


## 4.4 Labour-Market Slack Over Time

The ABS Labour Force data provide a long-run view of labour-market slack in Australia. Figure @fig-labour-market-slack presents monthly unemployment, underemployment and underutilisation rates from February 1978 to July 2026. In this analysis, underutilisation is used as the primary observable proxy for labour-market slack, while unemployment and underemployment are examined separately to distinguish the different components of unused labour capacity.

```{r}
#| label: fig-labour-market-slack
#| fig-cap: "Unemployment, underemployment and underutilisation in Australia"
#| warning: false
#| message: false

library(tidyverse)

slack <- read_csv(
  "../data/processed/labour_market_slack.csv",
  show_col_types = FALSE
)

slack_long <- slack |>
  select(
    date,
    unemployment_rate,
    underemployment_rate,
    underutilisation_rate
  ) |>
  pivot_longer(
    cols = c(
      unemployment_rate,
      underemployment_rate,
      underutilisation_rate
    ),
    names_to = "measure",
    values_to = "rate"
  ) |>
  mutate(
    measure = recode(
      measure,
      unemployment_rate = "Unemployment",
      underemployment_rate = "Underemployment",
      underutilisation_rate = "Underutilisation"
    )
  )

ggplot(slack_long, aes(x = date, y = rate, colour = measure)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = "Australian Labour-Market Slack Over Time",
    x = NULL,
    y = "Rate (%)",
    colour = "Measure"
  ) +
  theme_minimal()
```

The three measures display substantial variation over the study period and illustrate why unemployment alone may not capture the full extent of unused labour capacity. Unemployment measures people who are without work and actively seeking and available for work, whereas underemployment captures employed people who want and are available for additional hours. Underutilisation combines unemployment and underemployment and therefore provides a broader measure of labour underutilisation.

To compare labour-market conditions across different stages of the sample, the analysis divides the period into six broad intervals: 1978–1989, 1990–1994, 1995–2007, 2008–2019, 2020–2021 and 2022 onwards. Table @tbl-slack-periods summarises the average unemployment, underemployment and underutilisation rates for each period.

```{r}
#| label: tbl-slack-periods
#| tbl-cap: "Average labour-market slack measures by period"
#| warning: false
#| message: false

period_summary <- slack |>
  mutate(
    period = case_when(
      date >= as.Date("1978-02-01") & date <= as.Date("1989-12-01") ~ "1978–1989",
      date >= as.Date("1990-01-01") & date <= as.Date("1994-12-01") ~ "1990–1994",
      date >= as.Date("1995-01-01") & date <= as.Date("2007-12-01") ~ "1995–2007",
      date >= as.Date("2008-01-01") & date <= as.Date("2019-12-01") ~ "2008–2019",
      date >= as.Date("2020-01-01") & date <= as.Date("2021-12-01") ~ "2020–2021",
      date >= as.Date("2022-01-01") ~ "2022–present"
    )
  ) |>
  group_by(period) |>
  summarise(
    observations = n(),
    unemployment = mean(unemployment_rate, na.rm = TRUE),
    underemployment = mean(underemployment_rate, na.rm = TRUE),
    underutilisation = mean(underutilisation_rate, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  period_summary,
  digits = 2,
  col.names = c(
    "Period",
    "Observations",
    "Unemployment (%)",
    "Underemployment (%)",
    "Underutilisation (%)"
  )
)
```

The period averages show that the level and composition of labour-market slack have varied considerably across the sample. During 1978–1989, average unemployment was 7.37%, while underemployment averaged 3.36%, resulting in average underutilisation of 10.7%. Labour-market slack was higher during 1990–1994, when average unemployment increased to 9.58% and average underutilisation reached 15.7%. The period also recorded a peak unemployment rate of approximately 11.2% and a peak underutilisation rate of approximately 18.2%.

Between 1995 and 2007, average unemployment declined to 6.52%, while average underemployment increased to 6.60%. Average underutilisation was 13.1%. A similar pattern is evident between 2008 and 2019. Average unemployment declined further to 5.41%, while underemployment increased to 7.71%, leaving average underutilisation at approximately 13.1%. These results indicate that reductions in unemployment were not always accompanied by equivalent reductions in broader labour underutilisation.

The COVID-19 period provides a distinct episode of elevated labour underutilisation. Between 2020 and 2021, average unemployment was 5.80%, while average underemployment increased to 9.32%. Average underutilisation reached 15.1%, with a peak of approximately 20.3%. The relatively large increase in underemployment during this period demonstrates the importance of considering insufficient working hours when examining labour-market slack during periods of substantial economic disruption.

From 2022 onwards, average unemployment declined to 3.96%, while average underemployment remained at 6.19%. Average underutilisation was 10.2%. These averages indicate a lower level of measured labour underutilisation than in the preceding periods. However, underemployment remained a meaningful component of total underutilisation, reinforcing the limitation of using unemployment alone to represent unused labour capacity. This period is also shorter than the historical intervals considered above, so comparisons should be interpreted with this difference in sample length in mind.

Overall, the historical analysis demonstrates that labour-market slack is not captured fully by unemployment alone. Although periods of elevated slack have generally coincided with higher unemployment, underemployment has also contributed substantially to total labour underutilisation, particularly in more recent periods. Underutilisation therefore provides a broader observable measure for the subsequent analysis, while unemployment and underemployment are retained as separate measures to examine the composition of slack.


## 4.5 Labour-Market Tightness Over Time

Job vacancies alone do not account for the number of workers available in the labour market. The vacancy-to-unemployment ratio constructed in Section 3 provides an additional perspective by measuring the number of job vacancies relative to the number of unemployed persons.

```{r}
#| label: fig-tightness
#| fig-cap: "Labour-market tightness in Australia"
#| warning: false
#| message: false

tightness <- read_csv(
  "../data/processed/labour_market_tightness.csv",
  show_col_types = FALSE
) |>
  mutate(Quarter = as.Date(Quarter))

ggplot(tightness, aes(x = Quarter, y = Tightness)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Vacancies per unemployed person",
    title = "Labour-Market Tightness in Australia",
    subtitle = "ABS Job Vacancies / Unemployed Persons"
  ) +
  theme_minimal()
```

Figure @fig-tightness shows substantial variation in labour-market tightness over the study period. The ratio was particularly low around early 2020, indicating that there were relatively few job vacancies compared with the number of unemployed persons. Tightness then increased rapidly and reached its highest level in 2022.

Following the 2022 peak, labour-market tightness generally declined. By May 2026, the ratio was approximately 0.49, meaning that there were approximately 0.49 job vacancies for every unemployed person.

The tightness measure therefore provides additional information beyond the vacancy series alone. An increase in job vacancies does not necessarily represent the same labour-market conditions if the number of unemployed persons is also changing. Considering vacancies relative to unemployment allows labour demand to be interpreted in relation to the available unemployed workforce.

The five missing observations between August 2008 and August 2009 correspond to the period in which the ABS Job Vacancies Survey was suspended. These observations remain missing in the tightness series rather than being imputed [@ABS2026JobVacancies].


## 4.6 Labour-Market Tightness and Slack
To provide an initial view of the relationship between labour-market tightness and slack, the quarterly tightness measure is compared with unemployment, underemployment and underutilisation. Figure @fig-tightness-slack shows the relationship across the three measures.

```{r}
#| label: fig-tightness-slack
#| fig-cap: "Labour-market tightness and measures of labour-market slack"
#| warning: false
#| message: false

combined <- read_csv(
  "../data/processed/combined_tightness_slack.csv",
  show_col_types = FALSE
)

combined_long <- combined |>
  select(
    Tightness,
    unemployment_rate,
    underemployment_rate,
    underutilisation_rate
  ) |>
  pivot_longer(
    cols = c(
      unemployment_rate,
      underemployment_rate,
      underutilisation_rate
    ),
    names_to = "measure",
    values_to = "rate"
  ) |>
  mutate(
    measure = recode(
      measure,
      unemployment_rate = "Unemployment",
      underemployment_rate = "Underemployment",
      underutilisation_rate = "Underutilisation"
    )
  )

ggplot(combined_long, aes(x = Tightness, y = rate)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ measure, scales = "free_y") +
  labs(
    x = "Labour-market tightness",
    y = "Rate (%)",
    title = "Labour-Market Tightness and Labour-Market Slack",
    subtitle = "Quarterly comparison of three slack measures"
  ) +
  theme_minimal()
```

The three panels show a generally negative relationship between labour-market tightness and each measure of labour-market slack. Periods with more job vacancies relative to unemployed persons tend to coincide with lower unemployment, underemployment and underutilisation.

The relationship appears across all three measures, although its strength differs between them. This provides an initial descriptive indication that tighter labour-market conditions are associated with lower levels of unused labour capacity.

These plots are descriptive and do not establish a causal relationship. The relationship is examined more formally in the subsequent analysis using changes in the variables and statistical measures of association.

# 5. Methodology
This section describes the analytical approach used to examine the relationship between job vacancies and labour-market slack in Australia. The analysis is primarily descriptive and focuses on identifying how measures of labour demand and labour-market slack move together over time. Particular attention is given to changes in the variables because the research question asks to what extent changes in job vacancies are associated with changes in labour-market slack.

## 5.1 Analytical Framework
The analysis considers labour-market conditions from both the demand and slack sides of the labour market. Labour demand is represented by job vacancy measures, while labour-market slack is represented by unemployment, underemployment and underutilisation.

Two measures of labour demand are available in the project. The Jobs and Skills Australia Internet Vacancy Index (IVI) provides a monthly indicator of newly lodged online job advertisements, while ABS Job Vacancies provides a quarterly stock measure of vacancies available for immediate filling. As discussed in Section 3, these measures capture different concepts and are therefore not treated as interchangeable.

ABS Job Vacancies are also combined with unemployed persons to construct the vacancy-to-unemployment ratio:

$$
\text{Labour-Market Tightness}_t =
\frac{\text{Job Vacancies}_t}
{\text{Unemployed Persons}_t}
$$

A higher value indicates that there are more vacancies relative to the number of unemployed persons, while a lower value indicates fewer vacancies relative to unemployed persons.

The analytical framework therefore examines the labour market from three related perspectives: vacancy activity as an indicator of labour demand, unemployment, underemployment and underutilisation as measures of labour-market slack, and the vacancy-to-unemployment ratio as a complementary measure of labour-market tightness.

The analysis focuses on statistical association rather than causality. Observed relationships between vacancies, tightness and slack are therefore interpreted as evidence that the variables move together, rather than evidence that changes in one variable directly cause changes in another.


## 5.2 Data Alignment
The variables used in the analysis are reported at different frequencies. The IVI and ABS Labour Force measures are available monthly, whereas ABS Job Vacancies are available quarterly.

For analyses involving ABS Job Vacancies and labour-market tightness, the monthly Labour Force observations are aligned with the corresponding ABS Job Vacancies reference months: February, May, August and November. This produces a quarterly dataset in which vacancy and slack measures refer to corresponding periods.

The five unavailable ABS Job Vacancies observations between August 2008 and August 2009 are retained as missing values rather than being estimated or interpolated. Consequently, observations requiring Job Vacancies or labour-market tightness are also unavailable for these periods.

This alignment allows job vacancies, labour-market tightness and the three measures of labour-market slack to be compared on a consistent quarterly basis.


## 5.3 Levels and Changes
The analysis considers both the levels of the labour-market indicators and their changes over time.

Analysis in levels is used to examine whether periods of relatively high labour demand or labour-market tightness tend to coincide with relatively high or low levels of labour-market slack. This provides an initial description of the relationship between the variables.

Because the research question focuses on changes in job vacancies and labour-market slack, quarter-to-quarter changes are also considered in the subsequent analysis. For a variable $X_t$, the change between consecutive observations is defined as:

$$
\Delta X_t = X_t - X_{t-1}
$$

A positive value represents an increase from the previous observation, while a negative value represents a decrease. Examining changes allows the analysis to focus on movements in the labour-market indicators rather than only comparing their absolute levels.

## 5.4 Correlation Analysis
Pearson correlation coefficients are used as a descriptive measure of the direction and strength of the linear association between the labour-market variables.

The correlation coefficient ranges from -1 to 1. A positive coefficient indicates that two variables tend to move in the same direction, while a negative coefficient indicates that they tend to move in opposite directions. Values closer to zero indicate a weaker linear association.

Correlations are examined for the labour-market indicators in levels and for their changes over time. The three measures of labour-market slack are considered separately so that the relationship is not assessed using unemployment alone.

Particular care is required when interpreting the relationship between labour-market tightness and unemployment because unemployed persons are included in the denominator of the tightness measure. Consequently, part of the negative relationship between these two variables may arise mechanically from the construction of the ratio. Comparisons with underemployment and underutilisation provide additional evidence because these measures are not directly used in the calculation of the tightness ratio.

The correlations are interpreted as measures of association only. A strong correlation does not establish that changes in vacancies or labour-market tightness cause changes in labour-market slack.

## 5.5 Analysis of Changes

The analysis considers both the levels and changes of labour-market indicators. While analysis in levels provides information about whether periods of relatively high labour-market tightess tend to coincide with higher or lower levels of labour-market slack, the research question focuses specifically on chnages in job vacancies and changes in labour market slack. The analysis of changes therefore provides a closer match in main research question. 

For each labour market variable, the change between consecutive quarterly observations is calculated as:

$$
\Delta X_t = X_t - X_{t-1}
$$
where $X_t$ represents the value of the variable in the current quarter and $X_{t-1}$ represents its value in the previous quarter. A positive value therefore represents an increase relative to the previous quarter, while a negative value represents a decrease. 

This transformation is applied to labour market tightness and the three measures of labour market slack. The change in tightness is calculated as :

$$
\Delta T_t = T_t - T_{t-1}
$$
where $T_t$ represents the vacancy-to-unemployment ratio. Changes in unemployment, underemployment and underutilisation are calculated similarly:
$$
\Delta U_t = U_t - U_{t-1}
$$
$$
\Delta UE_t = UE_t - UE_{t-1}
$$
$$
\Delta UR_t = UR_t - UR_{t-1}
$$
where $U_t$ denotes the unemployment rate, $UE_t$ denotes the underemployment rate and $UR_t$ denotes the underutilisation rate.

The change-based analysis is used to examine whether periods in which labour-market tightness increases or decreases are associated with corresponding movements in labour-market slack. This reduces the emphasis on long-run differences in the level of the variables and instead focuses on quarter-to-quarter movements.
The analysis also compares changes in labour-market conditions between observations classified as high-tightness and high-slack periods and the remaining observations. This provides an exploratory way of identifying whether periods in which relatively high labour-market tightness coexists with relatively high labour-market slack exhibit different patterns of change.
The change analysis remains descriptive. A correlation between changes in two variables indicates that their movements are associated over the period examined, but it does not establish that a change in one variable causes a change in the other.


# References