---
title: "Labour-Market Tightness and Slack in Australia"
author: "Rimlan & Trisha"
format: html
bibliography: references.bib
---

**Executive Summary**

To be completed after the analyis and discussion are finalised.

# 1. Introduction 

## 1.1 Background
The labour market plays an important role in the Australian economy by connecting employers seeking workers with individuals who are available and willing to work. Labour-market conditions change over time as the demand for workers and the availability of labour respond to economic conditions. Understanding these changes requires consideration of both the demand for labour and the extent to which available labour resources are being utilised.

One indicator of labour demand is the number of job vacancies. A higher level of vacancies generally indicates greater demand from employers for additional workers. However, different vacancy measures capture different aspects of recruitment activity. The Australian Bureau of Statistics (ABS) measures job vacancies as jobs that are available for immediate filling on a specific reference date and for which recruitment action has been taken [@ABS2026JobVacancies]. As a result, ABS Job Vacancies represents a stock measure of vacancies at a particular point in time.

Jobs and Skills Australia (JSA) provides another measure of labour demand through the Internet Vacancy Index (IVI). The IVI is a monthly series based on newly lodged online job advertisements from SEEK, CareerOne and Workforce Australia. It provides detailed information by occupation, skill level and geographic area and can be used as an indicator of recruitment activity and labour demand. However, because it measures newly lodged advertisements during each month rather than vacancies existing at a particular point in time, it differs conceptually from the ABS Job Vacancies measure.

Labour demand alone does not provide a complete picture of labour-market conditions. The supply side of the labour market is also important, particularly the amount of labour that is available but not fully utilised. This is commonly described as labour-market slack or labour underutilisation. Unemployment is one measure of unused labour capacity, but it does not capture workers who already have employment but would prefer and are available to work additional hours. The ABS therefore also publishes measures of underemployment and underutilisation. Underutilisation combines unemployment and underemployment and provides a broader measure of unused labour capacity in the labour market.

Considering labour demand and labour-market slack together can provide a more complete view of labour-market conditions. For example, a period with relatively high demand for workers and limited available labour represents different conditions from a period with weak labour demand and substantial unused labour capacity. This motivates examining not only vacancies and slack separately, but also how they move in relation to one another over time.

A related concept is labour-market tightness, which compares labour demand with the availability of workers. In this project, an initial measure of tightness is constructed as the ratio of ABS Job Vacancies to unemployed persons. A higher vacancy-to-unemployment ratio represents more vacancies relative to unemployed workers, while a lower ratio represents fewer vacancies relative to the number of unemployed workers. This measure provides an additional perspective on the balance between labour demand and available labour.

## 1.2 Research Motivation
Job vacancies, unemployment, underemployment and underutilisation each provide useful information about the Australian labour market, but they measure different aspects of labour-market conditions. Examining any one indicator independently may therefore provide an incomplete picture.

This distinction is particularly relevant when measuring labour-market slack. Unemployment identifies people without employment who satisfy specific job-search and availability criteria, whereas underemployment captures employed people whose labour is not being fully utilised. The ABS defines underemployed workers as employed people who would prefer and are available to work more hours than they currently have. The underutilisation rate provides a broader indicator by combining unemployed and underemployed people relative to the labour force.

The measurement of labour demand also requires careful consideration. The IVI captures the flow of newly lodged online job advertisements, while ABS Job Vacancies measures the stock of vacancies available for immediate filling on a reference date. In addition, the IVI does not represent all recruitment activity in Australia because it only covers advertisements from its contributing online job boards and excludes recruitment through other websites and informal channels. These differences mean that the choice of vacancy measure depends on the purpose of the analysis.

This project is therefore motivated by the need to examine labour demand and labour-market slack jointly rather than treating them as independent indicators. In particular, examining changes in these measures allows the analysis to investigate whether increases or decreases in job vacancies are associated with corresponding changes in unused labour capacity.

Labour-market tightness is also considered as a complementary measure. By relating the stock of job vacancies to the number of unemployed persons, tightness provides information about vacancies relative to the available pool of unemployed workers. This allows the project to investigate both the direct relationship between changes in vacancies and slack and the broader relationship between labour-market tightness and different measures of slack.

## 1.3 Research Question
Based on this motivation, the main research question is:

To what extent are changes in job vacancies associated with changes in labour-market slack in Australia?

The wording of the research question focuses on association rather than causation. The purpose of the analysis is to determine whether movements in labour demand are systematically related to movements in labour-market slack and to examine how this relationship differs depending on the measure of slack considered.

The project considers three measures of labour-market slack: unemployment, underemployment and underutilisation. This allows the relationship to be examined using both a narrower measure of unused labour capacity and broader measures that account for workers who are employed but would prefer additional hours.

Labour-market tightness is examined as a complementary perspective rather than as a replacement for the main vacancy measure. This distinction is important because the research question specifically concerns changes in job vacancies, while tightness incorporates information from both vacancies and unemployment.

## 1.4 Research Objectives
The project has four main objectives:

To describe how job vacancies and labour-market slack have changed in Australia over the available data period.
To compare alternative measures of labour-market slack, including unemployment, underemployment and underutilisation.
To examine the association between changes in job vacancies and changes in labour-market slack, including potential lagged relationships.
To investigate labour-market tightness as an additional indicator of the balance between labour demand and available labour.

The analysis uses data from the Australian Bureau of Statistics and Jobs and Skills Australia. The IVI is used to examine monthly online recruitment activity, while ABS Job Vacancies and Labour Force data provide measures of vacancies and labour availability. The different data sources are examined carefully because their definitions, frequency and measurement concepts are not identical.

## 1.5 Report Structure
The remainder of this report is organised as follows. Section 2 introduces the main labour-market concepts and provides background on labour demand, labour-market slack and labour-market tightness. Section 3 describes the datasets, variable construction and measurement choices used in the analysis. Section 4 presents exploratory analysis of labour demand, slack and tightness in Australia. Section 5 describes the analytical methodology, including the analysis of changes, correlations and potential lagged relationships. Section 6 presents the empirical results, followed by a discussion of their interpretation in Section 7. Section 8 discusses the limitations of the analysis and possible extensions, and Section 9 concludes the report.

# 2. Labour-Market Concepts and Background

## 2.1 Labour Demand and Job Vacancies

## 2.2 Labour-Market Slack

## 2.3 Labour-Market Tightness

## 2.4 Relationship Between Labour Demand and Slack

## 2.5 Australian Labour-Market Context

# 3. Data and Measurement

## 3.1 Jobs and Skills Australia Internet Vacancy Index

### 3.1.1 Data Source and Collection

The Internet Vacancy Index (IVI), published by Jobs and Skills Australia
(JSA), is used in this project as a monthly indicator of labour demand in
Australia. The IVI measures online job advertisements newly lodged during
each reference month on three contributing job boards: SEEK, CareerOne and
Workforce Australia [@JSAIVIMethodology].

The IVI is based on administrative data provided by the contributing job
boards rather than data collected through an employer survey. JSA processes
the advertisements and classifies them by occupation, skill level and
geographical location. Duplicate advertisements posted on the same or
multiple contributing job boards are removed during this process
[@JSAIVIMethodology].

### 3.1.2 Measurement and Coverage
The IVI is available as a monthly time series from January 2006. The data
provide information by occupation, skill level, state and territory, and
regional areas. For the national labour-demand analysis in this project,
the seasonally adjusted Australian total series is used.

An important characteristic of the IVI is that it measures newly lodged
advertisements during a month. It therefore represents a flow of online
recruitment activity rather than the stock of vacancies that remain
unfilled at a particular point in time [@JSAIVIMethodology].

### 3.1.3 Limitations of the IVI
The IVI does not represent all vacancies in the Australian labour market.
It only includes advertisements from the contributing online job boards
and therefore excludes recruitment through other job boards, employer
websites, social media, newspapers and informal methods such as word of
mouth. In addition, one advertisement may represent multiple available
positions, while some vacancies may be filled without being advertised
online [@JSAIVIMethodology].

For these reasons, the IVI is interpreted in this project as an indicator
of online recruitment activity and labour demand rather than a complete
measure of all job vacancies in Australia.


## 3.2 ABS Job Vacancies

### 3.2.1 Job Vacancies Survey
The second measure of labour demand is obtained from the Australian Bureau
of Statistics (ABS) Job Vacancies Survey (JVS). Unlike the IVI, which is
based on online job advertisements, the JVS collects information directly
from a sample of employing organisations. Approximately 4,000 statistical
units are selected from the ABS Business Register, with data collected
through online electronic collection and/or telephone interviews
[@ABS2026JobVacancies].

The ABS defines a job vacancy as a job that is available for immediate
filling on the survey reference date and for which recruitment action has
been taken [@ABS2026JobVacancies].

### 3.2.2 Measurement and Coverage
The JVS is conducted quarterly. Its reference date is the third Friday of
the middle month of each quarter, corresponding to February, May, August
and November. The number of vacancies therefore refers to vacancies
existing on a particular reference date rather than vacancies accumulated
over the entire quarter [@ABS2026JobVacancies].

For this project, the seasonally adjusted Australian total job-vacancy
series is used. ABS classifies job vacancies as a stock measure because
they are measured at a specific point in time [@ABS2026JobVacancies].

There are five missing quarterly observations between August 2008 and
August 2009 because the Job Vacancies Survey was suspended after May 2008
and reinstated in November 2009. These observations are retained as
missing rather than being imputed, as the ABS does not provide official
estimates for the gap period [@ABS2026JobVacancies].

### 3.2.3 IVI versus ABS Job Vacancies
Although both series provide information about labour demand, they measure
different concepts. The IVI counts newly lodged online job advertisements
during a month and is therefore a flow measure of recruitment activity.
In contrast, ABS Job Vacancies measures the number of vacancies available
for immediate filling on a particular reference date and is therefore a
stock measure [@JSAIVIMethodology; @ABS2026JobVacancies].

This distinction is important for the analysis. The IVI is used to examine
monthly changes in recruitment demand, while ABS Job Vacancies is used in
constructing the quarterly labour-market tightness measure. Using ABS Job
Vacancies for the tightness measure provides a more consistent comparison
with the number of unemployed persons, since both represent quantities
measured for a particular reference period.


## 3.3 ABS Labour Force Data
This section describes the ABS Labour Force data used to measure
unemployment, underemployment and underutilisation.

[To be completed with the labour-market slack component.]


## 3.4 Labour-Market Tightness

### 3.4.1 Vacancy-to-Unemployment Ratio
Labour-market tightness is used to represent the balance between labour
demand and the availability of unemployed workers. In this project,
tightness is initially measured using the vacancy-to-unemployment ratio:

$$
\text{Labour-market tightness}_t =
\frac{\text{Job Vacancies}_t}
{\text{Unemployed Persons}_t}
$$

A higher value indicates that there are more job vacancies relative to
the number of unemployed persons, representing a tighter labour market.
Conversely, a lower value indicates fewer vacancies relative to unemployed
persons.

### 3.4.2 Construction of the Measure
The numerator is the seasonally adjusted Australian total ABS Job
Vacancies series, while the denominator is the seasonally adjusted number
of unemployed persons from the ABS Labour Force data.

Because Job Vacancies data are quarterly while Labour Force data are
monthly, unemployed-person observations are matched to the corresponding
vacancy reference months of February, May, August and November. Both
series are measured in thousands of persons or positions, so the units
cancel when the ratio is calculated.

The resulting measure provides a quarterly series of labour-market
tightness that can subsequently be compared with the measures of
labour-market slack used in this project.

## 4.1 ABS Job Vacancies Over Time
The ABS Job Vacancies series provides a quarterly view of labour demand in Australia. Figure @fig-job-vacancies shows the seasonally adjusted number of job vacancies over the study period.

```{r}
#| label: fig-job-vacancies
#| fig-cap: "Seasonally adjusted job vacancies in Australia"
#| warning: false
#| message: false

library(tidyverse)

job_vacancies <- read_csv(
  "../data/processed/abs_job_vacancies_clean.csv",
  show_col_types = FALSE
) |>
  mutate(Quarter = as.Date(Quarter))

ggplot(job_vacancies, aes(x = Quarter, y = Job_Vacancies)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Job vacancies ('000)",
    title = "Job Vacancies in Australia",
    subtitle = "Seasonally adjusted ABS Job Vacancies"
  ) +
  theme_minimal()
```

Job vacancies changed considerably over the study period. Vacancies declined during the late 2000s, although the series contains a gap between August 2008 and August 2009 due to the suspension of the Job Vacancies Survey during this period [@ABS2026JobVacancies].

A particularly large decline is visible in 2020, followed by a strong increase during 2021 and 2022. Job vacancies reached their highest levels around 2022 before declining over the following years.

These patterns indicate that labour demand has varied substantially over time. However, the figure is descriptive and does not by itself establish the causes of these movements.

## 4.2 Internet Vacancy Index Over Time
The Internet Vacancy Index provides a higher-frequency view of recruitment activity because it records newly lodged online job advertisements each month. Figure @fig-ivi-total shows the seasonally adjusted Australian total IVI.

```{r}
#| label: fig-ivi-total
#| fig-cap: "Internet Vacancy Index for Australia"
#| warning: false
#| message: false

library(tidyverse)

ivi_anzsco2 <- read_csv(
  "../data/processed/ivi_anzsco2_clean.csv",
  show_col_types = FALSE
)

ivi_australia <- ivi_anzsco2 |>
  filter(
    Level == 1,
    ANZSCO_CODE == "0",
    State == "AUST"
  ) |>
  arrange(Month)

ggplot(ivi_australia, aes(x = Month, y = Vacancies)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = NULL,
    y = "Online job advertisements",
    title = "Internet Vacancy Index in Australia",
    subtitle = "Seasonally adjusted national series"
  ) +
  theme_minimal()
```

The IVI also shows substantial variation in recruitment activity over time. Online job advertisements increased during the earlier part of the series before declining around 2008–2009. Recruitment activity subsequently recovered, although further periods of decline and recovery are visible.

One of the most pronounced movements occurs in 2020, when the IVI falls sharply. This is followed by a strong increase, with recruitment activity reaching particularly high levels around 2022. The index subsequently declines through the following years.

The broad movements are similar to those observed in ABS Job Vacancies. However, the two series should not be interpreted as identical measures. The IVI measures the flow of newly lodged online advertisements each month, whereas ABS Job Vacancies measures the stock of vacancies available for immediate filling at a reference date [@JSAIVIMethodology; @ABS2026JobVacancies].

# References