---
title: "Job Vacancies and Labour-Market Slack in Australia"
subtitle: "Examining the short-run relationship between labour demand and labour-market slack"
author:
  - "Trisha Andrea"
  - "Rimlan Alva"
date: today
toc: false
format:
  presentation-revealjs+letterbox: default
---

```{r}
#| label: load-packages
#| include: false
library(knitr)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
options(width = 70)

combined <- read_csv(
  "../data/processed/combined_tightness_slack.csv",
  show_col_types = FALSE
)

combined <- combined |>
  mutate(Quarter = as.Date(Quarter))
```
---
## Why does this matter?

::: columns

::: {.column width="55%"}

### Job vacancies signal labour demand

- More vacancies generally indicate stronger demand for workers.
- Labour-market slack shows how much available labour is not being fully used.
- Understanding how these move together can provide insight into changing labour-market conditions.

:::

::: {.column width="45%"}

### The key idea

**Higher labour demand**

⬇

**More job vacancies**

⬇

**Potentially less labour-market slack**

:::

:::

## Research Question

<br><br>

### To what extent are changes in **job vacancies** associated with short-run changes in **labour-market slack** in Australia?

## Data

We combine two main sources of Australian labour-market data.

::: columns

::: {.column width="48%"}

### Job Vacancies

**Internet Vacancy Index (IVI)**  
Jobs and Skills Australia

- Measures online job advertisements
- Used as our measure of labour demand

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}

### Labour-Market Slack

**Labour Force data**  
Australian Bureau of Statistics (ABS)

- Unemployment
- Underemployment
- Labour underutilisation

:::

:::

## Methodology

### 1. Calculate short-run changes

**Quarter-to-quarter:**

Δ Job Vacancies  
Δ Unemployment · Δ Underemployment · Δ Underutilisation

<br>

### 2. Compare vacancy and slack changes

**Δ Vacancies ↔ Δ Slack**

<br>

### 3. Examine the relationship

**Scatter plots · Linear trends · Correlations**

## Labour-Market Tightness Over Time

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5

ggplot(
  combined,
  aes(
    x = Quarter,
    y = Tightness
  )
) +
geom_line(
  linewidth = 0.9,
  na.rm = TRUE
) +

annotate(
  "text",
  x = as.Date("2009-03-01"),
  y = 0.43,
  label = "Job vacancy data\nunavailable",
  size = 3.5,
  fontface = "italic"
) +

labs(
  subtitle = "Job vacancies per unemployed person",
  x = NULL,
  y = "Vacancies per unemployed person",
  caption = "Source: Australian Bureau of Statistics"
) +
  scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    panel.grid.minor = element_blank()
  )
```

## Tightness and Labour-Market Slack

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 11
#| fig-height: 5

combined_long <- combined |>
  select(
    Quarter,
    Tightness,
    unemployment_rate,
    underemployment_rate,
    underutilisation_rate
  ) |>
  pivot_longer(
    cols = c(
      unemployment_rate,
      underemployment_rate,
      underutilisation_rate
    ),
    names_to = "measure",
    values_to = "slack_rate"
  ) |>
  mutate(
    measure = recode(
      measure,
      unemployment_rate = "Unemployment",
      underemployment_rate = "Underemployment",
      underutilisation_rate = "Underutilisation"
    )
  )

ggplot(
  combined_long,
  aes(
    x = Tightness,
    y = slack_rate
  )
) +
  geom_point(
    alpha = 0.6,
    na.rm = TRUE
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    na.rm = TRUE
  ) +
  facet_wrap(
    ~ measure,
    scales = "free_y"
  ) +
  labs(
    x = "Vacancies per unemployed person",
    y = "Slack rate (%)",
    caption = "Source: Australian Bureau of Statistics"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(
      face = "bold"
    ),
    panel.grid.minor = element_blank()
  )
```

## Interactive Labour-Market Dashboard

We developed a **Shiny dashboard** to make the analysis interactive and easier to explore.

::: columns

::: {.column width="50%"}

### Explore

- Labour demand and job vacancies
- Labour-market slack and tightness
- Vacancy–slack relationships
- Wage and occupation patterns

:::

::: {.column width="50%"}

### Interact

- Filter time periods and indicators
- Compare different slack measures
- Explore interactive plots and tables
- Search and download project data

:::

:::

## Main Results

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 11
#| fig-height: 5

changes_long <- combined |>
  arrange(Quarter) |>
  mutate(
    change_vacancies = Job_Vacancies - lag(Job_Vacancies),
    change_unemployment = unemployment_rate - lag(unemployment_rate),
    change_underemployment = underemployment_rate - lag(underemployment_rate),
    change_underutilisation = underutilisation_rate - lag(underutilisation_rate)
  ) |>
  select(
    Quarter,
    change_vacancies,
    change_unemployment,
    change_underemployment,
    change_underutilisation
  ) |>
  pivot_longer(
    cols = c(
      change_unemployment,
      change_underemployment,
      change_underutilisation
    ),
    names_to = "measure",
    values_to = "change_slack"
  ) |>
  mutate(
    measure = recode(
      measure,
      change_unemployment = "Unemployment",
      change_underemployment = "Underemployment",
      change_underutilisation = "Underutilisation"
    )
  ) |>
  filter(
    !is.na(change_vacancies),
    !is.na(change_slack)
  )

ggplot(
  changes_long,
  aes(
    x = change_vacancies,
    y = change_slack
  )
) +
  geom_point(
    alpha = 0.6
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  facet_wrap(
    ~ measure,
    scales = "free_y"
  ) +
  labs(
    x = "Change in job vacancies",
    y = "Change in labour-market slack",
    caption = "Source: Australian Bureau of Statistics"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```
## Interpretation & Limitations

### Interpretation

**↑ Job vacancies → ↓ Labour-market slack**

The negative association appears across all three slack measures.

<br>

### Limitations

- **Association does not imply causation**
- Economic shocks may affect both vacancies and slack
- Some periods have missing job-vacancy data

## Conclusion

### What did we find?

Changes in job vacancies are **negatively associated** with short-run changes in labour-market slack.

The pattern is observed across:

**Unemployment · Underemployment · Underutilisation**

<br>

### Key takeaway

**Stronger labour demand tends to coincide with lower labour-market slack.**

## Thank You {.center}
### Any questions?